### 2023-04-18 paper.Rmd
more work on formalization

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ will lead to similar variances across the $K$ groups in the $P$ attributes (**only if the group means are close to the overall mean, discuss this as suggested in reviewer 1's comment 4**; $\mathit{SSE_{Var}}$ is only valid if the group means are close to the overall means of the data.).

@@

As discussed (**TODO**) in the previous section, an exclusive maximization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is not feasible without also considering the standard $\mathit{SSE}$ (i.e., equalizing means). Moreover, only equalizing variances between groups is most likely not even of interest in most applications. Hence, when striving for similarity in group variances by maximizing $\mathit{SSE_{Var}}$, it is at the same time necessary and worthwhile to strive for similarity in means, thus to maximize $\mathit{SSE}$ at the same time. Hence, a combination of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ should be employed to simultaneously optimize both criteria, which also represents a reasonable idea when striving for overall between-group similarity. The most simple approach for such a multicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, all criteria are computed independently from each other and are then combined into a single criterion through a weighted sum

$$
\mathit{SSE_{kplus}} = 
w_1 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{x}_j ^{(k)})^2 + 
w_2 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{y}_j ^{(k)})^2
$$
where the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively. Maximizing $\mathit{SSE_{kplus}}$ for a given $K$ will lead to similar means as well as variances between groups.

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria at the same time if the underlying objectives are in conflict with each other [@ferligoj1992direct]. It is therefore unclear whether optimizing a bicriterion objective such as $\mathit{SSE_{kplus}}$ can yield a partitioning where both constituting criteria (mean and variance) are similar between groups, because it is unknown if these objectives tend to align with each other or are in conflict. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ (i.e., $w_1$ and $w_2$ were set to 1 when computing $\mathit{SSE_{kplus}}$), which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree, indicating that these objectives tend to align and are not in conflict with each other.

@@

## Including higher order moments

Note that the k-plus logic to equalizing group variances as well as group means can be seamlessly generalized to higher order moments. The *t*'th sample moment of a variable $x = (x_1, \ldots, x_N)$ can be computed as

@@

where $t = 2$ gives the variance, $t = 3$ the skewness, and $t = 4$ the kurtosis [@joanes1998]. Thus, by changing $t$---the power of the deviation between mean and data points---we can use the k-plus logic to equalize any desired sample moment between groups.

Next, I provide some notation needed to compute the general $\mathit{SSE}$ for k-plus anticlustering, which may include arbitrary distribution moments as part of the computation. For each moment $t$ and for column of the original data matrix $x_{.j} =  (x_{1j}, \ldots, x_{Nj})$ ($j = 1, \ldots, P$) with mean $\overline{\mathbf{x}}_j$, a new variable $x_{t.j} = (x_{t1j}, \ldots, x_{tNj})$ is computed where

$$
    x_{tij} = 
\begin{cases}
    x_{ij} & \text{if } t = 1\\
    (x_{ij} - \overline{\mathbf{x}}_j)^t, & \text{if } t > 1.
\end{cases}
$$

This computation results in an augmented data matrix $\mathbf{\hat{X}}_{N \times (T \cdot P)} = \{x_{tij}\}_{N \times (T \cdot P)}$ where $T$ is the number of distribution moments that should be equalized between groups. Note that when defining $x_{tij}$, the condition $t = 1$ is included to keep the unchanged original data as the first $P$ columns in the data, which is used to equalize the the mean---the first moment---between groups. Using the thus defined data matrix $\mathbf{\hat{X}}_{N \times (T \cdot P)}$, the generalized multicriterion for k-plus anticlustering can be formalized by the formula

$$
\mathit{SSE_{kplusGeneralized}} =
  \sum\limits_{t=1}^{T} 
  w_t \left(
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{tij} - \overline{x}_{.j} ^{(k)})^2
  \right)
$$

where $w_t$ is the weight associated with the $t'th$ moment and $\overline{x}_{.j} ^{(k)}$ is the mean of the $j'th$ variable in the $k'th$ group. Maximizing $\mathit{SSE_{kplusGeneralized}}$ with a given number of moments $T$ can be used to equalize these distribution moments between $K$ groups. Note that according to this definition the inclusion of a distribution moment $t$ always implies that all moments $z < t$ are also included as part of the criterion $\mathit{SSE_{kplusGeneralized}}$; it is not deemed worthwhile to exclude lower order moments in favor of higher order moments. In particular, the first moment, represented by the original data, should always be included because otherwise the criterion $\mathit{SSE_{kplusGeneralized}}$ is not valid for higher order moments.

>>>>

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ can be used to create groups that have similar variances across the $P$ attributes. As a caveat, it is crucial to note that maximizing $\mathit{SSE_{Var}}$ only reliably yields similar group variances if the group centroids $\overline{\mathbf{x}}^{(k)}$, ($k = 1, \dots, K$) are close to the overall data centroid $\overline{\mathbf{x}}$. To reiterate, this condition has to be met for the original data $\mathbf{X}_{N \times P}$. Thus, only considering $\mathbf{Y}_{N \times P}$ in order to minimize discrepancies in variance may not yield satisfying results; instead, it has to be ensured that the group means of the original data are also similar between groups. Fortunately, minimizing discrepancy with regard to group means can be accomplished by standard k-means anticlustering, i.e., the maximization of $\mathit{SSE}$. Combining the maximization of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ at the same time leads to the k-plus approach for anticlustering, which is detailed in the next section.

@@

As discussed in the previous section, an exclusive maximization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is not feasible without also considering the standard $\mathit{SSE}$ (i.e., equalizing means). Moreover, an exclusive maximizing of $\mathit{SSE_{Var}}$ is most likely not even interesting in any real application. Hence, when striving for similarity in group variances by maximizing $\mathit{SSE_{Var}}$, it is at the same time necessary and worthwhile to strive for similarity in means, and thus to maximize $\mathit{SSE}$ at the same time. The most simple approach for a multicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, all criteria are computed independently from each other and are then combined into a single criterion through a weighted sum. By combining $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ in a single criterion, we define the objective function for k-plus anticlustering:

$$
\mathit{SSE_{kplus}} = 
w_1 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{x}_j ^{(k)})^2 + 
w_2 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{y}_j ^{(k)})^2
$$
Maximizing $\mathit{SSE_{kplus}}$ will lead to similar means and variances between $K$ groups, while the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively.

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict with each other [@ferligoj1992direct]. As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown whether maximizing it can yield a partitioning where both constituting criteria (mean and variance) are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ (i.e., $w_1$ and $w_2$ were set to 1 when computing $\mathit{SSE_{kplus}}$), which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree, indicating that these objectives tend to align and are not in conflict with each other. Moreover, the scatterplot in Figure 1 shows that there is a high concentration of partitions in the upper-right quadrant where the two objectives are in high accordance.

@@

## Generalizing k-plus anticlustering to higher order moments

The k-plus logic to equalizing group variances as well as group means can be seamlessly generalized to higher order moments. The *t*'th sample moment of a variable $x = (x_1, \ldots, x_N)$ can be computed as

@@

where $t = 2$ gives the variance, $t = 3$ the skewness, and $t = 4$ the kurtosis [@joanes1998]. Thus, by changing $t$---the power of the deviation between mean and data points---we can use the k-plus logic to equalize any desired sample moment between groups. That is, we compute new variables for each original variable and for each distribution moment that should be considered in the anticlustering process. To formalize this idea, I now provide some notation to compute a generalized $\mathit{SSE}$ for k-plus anticlustering, which can include arbitrary distribution moments as part of the criterion. First, we define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{.j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean is $\overline{\mathbf{x}}_j$. Let $T$ be the number of distribution moments that should be equalized between $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{.j}$, a new *t-moment variable* $\hat{x}_{t.j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

$$
    \hat{x}_{tij} = 
\begin{cases}
    x_{ij} & \text{if } t = 1\\
    (x_{ij} - \overline{\mathbf{x}}_j)^t, & \text{if } t > 1.
\end{cases}
$$

This computation results in an augmented data matrix $\mathbf{\hat{X}}_{N \times \mathit{TP}} = \{\hat{x}_{tij}\}_{N \times \mathit{TP}}$, i.e.

$$
  \mathbf{\hat{X}}_{N \times \mathit{TP}} =
  \left[ {\begin{array}{ccc|ccc|c|ccc}
    x_{11} & \cdots & x_{1P} & 
        (x_{11} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{1P} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{11} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{1P} - \overline{\mathbf{x}}_P)^T \\
    x_{21} & \cdots & x_{2P} & 
        (x_{21} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{2P} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{21} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{2P} - \overline{\mathbf{x}}_P)^T \\
    \vdots & \ddots & \vdots & \vdots & \ddots & \vdots & \cdots & \vdots & \ddots & \vdots\\
    x_{N1} & \cdots & x_{NP} & 
        (x_{N1} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{NP} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{N1} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{NP} - \overline{\mathbf{x}}_P)^T \\
  \end{array} } \right]
$$

Note that when defining $\hat{x}_{tij}$, the condition $t = 1$ is separately included to keep the unchanged original data as the first $P$ columns in the data, which is used to equalize the the mean---the first moment---between groups.

Using the data matrix $\mathbf{\hat{X}}_{N \times \mathit{TP}}$ and a vector of weights $\mathbf{w} = (w_1, \dots, w_T$), the generalized multicriterion for k-plus anticlustering $\mathit{SSE_{kplusGeneralized}}$ can be formalized:

$$
\mathit{SSE_{kplusGeneralized}} =
  \sum\limits_{t=1}^{T} 
  w_t \left(
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{tij} - \overline{x}_{tj} ^{(k)})^2
  \right)
$$

Here, $w_t$ is the weight associated with the $t$'th moment, $\overline{x}_{tj} ^{(k)}$ is mean for the $k$'th group on the $j$'th $t$-moment variable. Maximizing $\mathit{SSE_{kplusGeneralized}}$ can be used to equalize the first $T$ distribution moments between $K$ groups.
